---
title: "Quickstart notebook"
output:
  html_document:
    self_contained: true
---

```{r setup, message=FALSE}
library(VGAM)
library(MASS)

set.seed(1)
```

# Purpose

This notebook provides a fast-running illustration of the repository workflow. It directly uses a reduced version of the same Gibbs-sampler code path as [`Section6.1/Table1/linearregression_gibbs_unbounded_fullprior.R`](Section6.1/Table1/linearregression_gibbs_unbounded_fullprior.R), so it is a genuine small-scale example for Table 1. The other results in `Table2`, `Table3`, and `Section6.2` are implemented with different models and code, so this notebook is not intended to reproduce those scripts directly. Instead, it gives a compact example of the same general pattern used throughout the repository: simulate data or use real data, compute summaries, add noise to achieve differential privacy, run the proposed algorithm, generate MCMC chains, and summarize the resulting outputs.

In this notebook we run only a single reduced unbounded example:

- one dataset,
- `eps = 1`,
- `ep_N = 0.5`,
- `num_cycles = 200`.

Ideally, `num_cycles` would be set to `10000`, but in this notebook we use `200` so the example remains fast enough for a quick check of the core unbounded-DP logic, while the section-specific scripts remain the source for reproducing the full results.

# Reduced Table 1 Gibbs sampler

The code chunk below defines the main functions used in the reduced Table 1 example, including data generation, privatized summary calculations, Gibbs updates, and the routine that runs the MCMC chain.

```{r}
###   the model is as follows:
###   x | mu, phi ~ N_p(mu, phi^{-1})
###   y | x, beta, tau ~ N(x*beta, tau^{-1} I)
###   with priors
###   mu ~ N(theta, Sigma)
###   Phi ~ Wishart_p(d,W)
###   beta | tau ~ N_{p+1}(m,tau^{-1}V^{-1})
###   tau ~ Gamma(a/2, b/2), using the shape, rate parameterization
###   In total, the parameters are (beta, tau, mu, Phi)
###   the hyperparameters are (m, V, a, b, theta, Sigma, d, W)
###   and sufficient statistics are (XtX, XtY, YtY, X_bar, and S=sum((X-X_bar)%*%top(X-X_bar)) )

## clamping + location-scale transformation
## use this function to normalize each covariate x[-,j] and y
normalize <- function(vec, left, right){
  return(2 * (pmin(right, pmax(left, vec)) - left) / (right - left) - 1 )
}

## prior on N
g_log <- function(N){
  return(0)
}

#log-probability of transitioning from N to N_.
q_log <- function(N_, N){
  if(N == 1 & N_ == 2){
    return(0)
  }
  if(N > 1 & abs(N - N_) == 1){
    return(log(1/2))
  }
  else{
    print("q is zero")
    return(-Inf)
  }
}

## initialize the Gibbs sampler----
rinit_data <- function(N, params){
  x = mvrnorm(n = N, mu = params$mu, Sigma = params$phi_inv)
  X = cbind(1, x)
  Y <- as.numeric(X %*% params$beta + rnorm(N, 0, s = sqrt(1 / params$tau)))
  return(data = list(N = N, X = X, Y = Y))
}

## initialize the Gibbs sampler----
rinit_gibbs <- function(N_guess, hyperparams = hyperparams, params = NULL){
  N_upper = 2 * N_guess
  if(is.null(params)){
    params = list()
    params$tau = hyperparams$a / hyperparams$b
    params$beta <- hyperparams$m
    params$mu <- hyperparams$theta
    params$phi <- ginv(hyperparams$W_inv)
    params$phi_inv <- ginv(params$phi)
  }
  state = rinit_data(N = N_upper, params)
  state$N = N_guess
  state$N_upper = N_upper
  state$params = params
  state$accMean_change = 0
  state$accMean_jump = 0
  return(state)
}

## compute distance from summary statistics to sdp--------
get_diffT <- function(state, clbounds, sdp){
  X_cl <- apply(state$X[1:state$N, ], 2, function(x) normalize(x, clbounds[1], clbounds[3]))
  X_cl[, 1] <- 1
  Y_cl <- sapply(state$Y[1:state$N], function(y) normalize(y, clbounds[2], clbounds[4]))
  if(is.null(sdp)){
    state$T1 <- t(X_cl) %*% Y_cl
    state$T2 <- t(Y_cl) %*% Y_cl
    state$T3 <- t(X_cl) %*% X_cl
  }
  else{
    state$T1 <- t(X_cl) %*% Y_cl - sdp[[1]]
    state$T2 <- t(Y_cl) %*% Y_cl - sdp[[2]]
    state$T3 <- t(X_cl) %*% X_cl - sdp[[3]]
  }
  return(state)
}

### these are not used in the DP density, just for the non-private posterior.
get_suffStats = function(state){
  state$XtX <- t(state$X[1:state$N, ]) %*% state$X[1:state$N, ]
  state$XtY <- t(state$X[1:state$N, ]) %*% state$Y[1:state$N]
  state$YtY <- t(state$Y[1:state$N]) %*% state$Y[1:state$N]
  state$Xbar <- state$XtX[1, 2:(p + 1)] / state$N
  state$S <- (state$N - 1) * cov(state$X[1:state$N, 2:(p + 1)])
  return(state)
}

## compute density of privacy noise---------
get_logeta <- function(state, N_DP, privacy_params){
  noise_scale = privacy_params$deltaa / privacy_params$eps
  return(sum(VGAM::dlaplace(state$T1, 0, noise_scale, TRUE)) +
    VGAM::dlaplace(state$T2, 0, noise_scale, TRUE) +
    sum(VGAM::dlaplace(state$T3[lower.tri(state$T3)], 0, noise_scale, TRUE)) +
    sum(VGAM::dlaplace(diag(state$T3)[2:length(state$T2)], 0, noise_scale, TRUE)) +
    VGAM::dlaplace(state$N - N_DP, 0, 1 / privacy_params$ep_N, TRUE))
}

## calculate individual contribution to t(x,y)
get_diffT_individual <- function(x, y, clbounds){
  xcl <- normalize(x, clbounds[1], clbounds[3])
  xcl[1] <- 1
  ycl <- normalize(y, clbounds[2], clbounds[4])
  diffti = list(diffti1 = xcl %o% ycl,
                diffti2 = t(ycl) %*% ycl,
                diffti3 = xcl %o% xcl)
  return(diffti)
}

## difference in each individual contribution to t(x,y)
get_diffeacht <- function(x, x_, y, y_, clbounds){
  xcl <- normalize(x, clbounds[1], clbounds[3])
  xcl[1] <- 1
  ycl <- normalize(y, clbounds[2], clbounds[4])
  xcl_ <- normalize(x_, clbounds[1], clbounds[3])
  xcl_[1] <- 1
  ycl_ <- normalize(y_, clbounds[2], clbounds[4])
  diffti = list(diffti1 = xcl_ %o% ycl_ - xcl %o% ycl,
                diffti2 = t(ycl_) %*% ycl_ - t(ycl) %*% ycl,
                diffti3 = xcl_ %o% xcl_ - xcl %o% xcl)
  return(diffti)
}

update_XiYi_gibbs_change <- function(i, state, hyperparams = hyperparams, privacy_params, N_DP){
  acc <- 0
  xi <- c(1, as.numeric(mvrnorm(n = 1, mu = state$params$mu, Sigma = state$params$phi_inv)))
  yi <- xi %*% state$params$beta + sqrt(1 / state$params$tau) * rnorm(1)
  diffti <- get_diffeacht(state$X[i, ], xi, state$Y[i], yi, privacy_params$clbounds)

  state_proposed = state
  state_proposed$X[i, ] <- xi
  state_proposed$Y[i] <- yi
  state_proposed$T1 = state$T1 + diffti$diffti1
  state_proposed$T2 = state$T2 + diffti$diffti2
  state_proposed$T3 = state$T3 + diffti$diffti3
  state_proposed$logeta = get_logeta(state_proposed, N_DP, privacy_params)

  logu <- log(runif(1))
  if(logu < state_proposed$logeta - state$logeta){
    acc <- 1
    state = state_proposed
  }
  state$acc <- acc
  return(state)
}

update_XiYi_gibbs_add <- function(state, hyperparams = hyperparams, privacy_params, N_DP){
  acc <- 0
  xi <- c(1, as.numeric(mvrnorm(n = 1, mu = state$params$mu, Sigma = state$params$phi_inv)))
  yi <- xi %*% state$params$beta + sqrt(1 / state$params$tau) * rnorm(1)
  diffti <- get_diffT_individual(xi, yi, privacy_params$clbounds)
  if(state$N + 1 > state$N_upper){
    print("augmented data structure doubled")
    state$X <- rbind(state$X, state$X)
    state$Y <- c(state$Y, state$Y)
    state$N_upper = 2 * state$N_upper
  }
  state_proposed = state
  state_proposed$X[state$N + 1, ] <- xi
  state_proposed$Y[state$N + 1] <- yi
  state_proposed$T1 = state$T1 + diffti$diffti1
  state_proposed$T2 = state$T2 + diffti$diffti2
  state_proposed$T3 = state$T3 + diffti$diffti3
  state_proposed$N = state$N + 1
  state_proposed$logeta = get_logeta(state_proposed, N_DP, privacy_params)

  logu <- log(runif(1))
  log_accept = state_proposed$logeta - state$logeta +
    q_log(state$N, state$N + 1) - q_log(state$N + 1, state$N) +
    g_log(state$N + 1) - g_log(state$N)

  if(logu < log_accept){
    acc <- 1
    state = state_proposed
  }
  state$acc <- acc
  return(state)
}

update_XiYi_gibbs_delete <- function(i, state, privacy_params, N_DP){
  acc <- 0
  diffti <- get_diffT_individual(state$X[i, ], state$Y[i], privacy_params$clbounds)
  state_proposed = state
  state_proposed$X[i, ] <- state$X[state$N, ]
  state_proposed$Y[i] <- state$Y[state$N]
  state_proposed$T1 = state$T1 - diffti$diffti1
  state_proposed$T2 = state$T2 - diffti$diffti2
  state_proposed$T3 = state$T3 - diffti$diffti3
  state_proposed$N = state$N - 1
  state_proposed$logeta = get_logeta(state_proposed, N_DP, privacy_params)

  logu <- log(runif(1))
  log_accept = state_proposed$logeta - state$logeta +
    q_log(state$N, state$N - 1) - q_log(state$N - 1, state$N) +
    g_log(state$N - 1) - g_log(state$N)
  if(logu < log_accept){
    acc <- 1
    state = state_proposed
  }
  state$acc <- acc
  return(state)
}

update_params_gibbs <- function(state, hyperparams = hyperparams){
  state = get_suffStats(state)
  m_star = ginv(state$XtX + hyperparams$V_inv) %*% (state$XtY + hyperparams$V_inv %*% hyperparams$m)
  V_inv_star = state$XtX + hyperparams$V_inv
  V_star = ginv(V_inv_star)
  a_star = hyperparams$a + state$N
  b_star = hyperparams$b + state$YtY + t(hyperparams$m) %*% hyperparams$V_inv %*% hyperparams$m - t(m_star) %*% V_inv_star %*% m_star

  tau_new = rgamma(n = 1, shape = a_star/2, rate = b_star/2)
  beta_new = mvrnorm(n = 1, mu = m_star, Sigma = V_star/tau_new)

  Sigma_star = ginv(hyperparams$Sigma_inv + state$N * state$params$phi)
  theta_star = Sigma_star %*% (hyperparams$Sigma_inv %*% hyperparams$theta + state$N * state$params$phi %*% state$Xbar)

  mu_new = mvrnorm(n = 1, mu = theta_star, Sigma = Sigma_star)

  d_star = hyperparams$d + state$N
  W_star = ginv(hyperparams$W_inv + state$S + state$N * (mu_new - state$Xbar) %*% t(mu_new - state$Xbar))

  phi_new = rWishart(n = 1, df = d_star, Sigma = W_star)[,, 1]

  params_new = list(tau = tau_new, beta = beta_new, mu = mu_new, phi = phi_new, phi_inv = ginv(phi_new))
  state$params = params_new
  return(state)
}

onestep_gibbs <- function(state, hyperparams, privacy_params, N_DP, swaps = 1, jumps = 1, fullUpdate = TRUE){
  if(is.nan(privacy_params$ep_N) == TRUE | privacy_params$ep_N == Inf)
    jumps = 0
  acc_change <- 0
  acc_jump <- 0
  if(fullUpdate == TRUE){
    for(i in 1:state$N){
      state <- update_XiYi_gibbs_change(i, state, hyperparams, privacy_params, N_DP)
      acc_change <- state$acc + acc_change
    }
    state$accMean_change <- acc_change / state$N
  }
  else if(swaps > 0){
    for(i in 1:swaps){
      i = sample(size = 1, 1:state$N)
      state <- update_XiYi_gibbs_change(i, state, hyperparams, privacy_params, N_DP)
      acc_change <- state$acc + acc_change
    }
    state$accMean_change <- acc_change / swaps
  }

  if(jumps > 0){
    for(index in 1:jumps){
      if(state$N == 1){
        state <- update_XiYi_gibbs_add(state, hyperparams, privacy_params, N_DP)
      }
      else{
        coin <- rbinom(n = 1, size = 1, prob = 1/2)
        if(coin == 1){
          state <- update_XiYi_gibbs_add(state, hyperparams, privacy_params, N_DP)
        }
        else if(coin == 0){
          i = sample(size = 1, 1:state$N)
          state <- update_XiYi_gibbs_delete(i, state, privacy_params, N_DP)
        }
        else{
          print("coin is broken")
        }
      }
      acc_jump <- state$acc + acc_jump
    }
  }
  state$accMean_jump <- acc_jump / jumps
  state = get_suffStats(state)
  state <- update_params_gibbs(state, hyperparams)
  return(state)
}

run_chain = function(iteration = 1, num_cycles = 15000, N = 1000, ep_N = NaN, eps, N_guess,
                     hyperparams = hyperparams, privacy_params = privacy_params,
                     swaps = 1, jumps = 1, fullUpdate = TRUE){

  params_true = list(beta = c(0, -1, 1),
                     tau = 1,
                     mu = c(1, -1),
                     phi = diag(p),
                     phi_inv = diag(p))

  set.seed(0)
  data = rinit_data(N, params_true)
  set.seed(iteration)

  N_DP = N + rlaplace(n = 1, location = 0, scale = 1 / privacy_params$ep_N)

  data = get_diffT(data, privacy_params$clbounds, sdp = NULL)
  noise_scale = privacy_params$deltaa / privacy_params$eps
  l1 <- matrix(VGAM::rlaplace(p + 1, 0, noise_scale), ncol = 1)
  l2 <- VGAM::rlaplace(1, 0, noise_scale)
  l3 <- matrix(VGAM::rlaplace((p + 1) * (p + 1), 0, noise_scale), ncol = (p + 1))
  l3[lower.tri(l3)] <- t(l3)[lower.tri(l3)]
  l3[1, 1] <- 0
  sdp = list(data$T1 + l1, data$T2 + l2, data$T3 + l3)

  if(is.na(N_guess))
    N_guess = max(2, ceiling(N_DP))
  state <- rinit_gibbs(N_guess, hyperparams)
  state <- get_diffT(state, privacy_params$clbounds, sdp)
  state$logeta <- get_logeta(state, N_DP, privacy_params)

  logeta_chain <- double(num_cycles)
  beta_chain <- matrix(NA, nrow = p + 1, ncol = num_cycles)
  tau_chain <- double(num_cycles)
  mu_chain <- matrix(NA, nrow = p, ncol = num_cycles)
  phi_chain <- matrix(NA, nrow = p^2, ncol = num_cycles)
  n_chain <- rep(NA, num_cycles)
  acc_chain_change <- double(num_cycles)
  acc_chain_jump <- double(num_cycles)

  for(iter in 1:num_cycles){
    state <- onestep_gibbs(state, hyperparams, privacy_params, N_DP, jumps = 1, fullUpdate = TRUE)
    logeta_chain[iter] <- state$logeta

    beta_chain[, iter] <- state$params$beta
    tau_chain[iter] <- state$params$tau
    mu_chain[, iter] <- state$params$mu
    phi_chain[, iter] <- as.numeric(state$params$phi)
    n_chain[iter] <- state$N
    acc_chain_jump[iter] <- state$accMean_jump
    acc_chain_change[iter] <- state$accMean_change
    if(iter %% 100 == 0) cat("iteration", iter, "(", 100 * iter / num_cycles, "% )\n")
  }

  outputs = list(logeta_chain = logeta_chain,
                 tau_chain = tau_chain,
                 beta_chain = beta_chain,
                 mu_chain = mu_chain,
                 phi_chain = phi_chain,
                 n_chain = n_chain,
                 acc_chain_jump = acc_chain_jump,
                 acc_chain_change = acc_chain_change,
                 sdp = sdp,
                 N_DP = N_DP)
  return(outputs)
}

num_cycles = 200
N = 1000
p = 2
eps = 1
ep_N = 0.5
hyperparams = list(p = 2,
                   m = rep(0, p + 1),
                   V_inv = diag(p + 1),
                   a = 2,
                   b = 2,
                   theta = rep(0, p),
                   Sigma_inv = diag(p),
                   d = 2,
                   W_inv = diag(p))

privacy_params = list(eps = eps,
                      ep_N = ep_N,
                      deltaa = p^2 / 2 + (2.5) * p + 2,
                      clbounds = c(-5, -5, 5, 5))

chain = run_chain(iteration = 123, num_cycles = num_cycles, N = N, N_guess = NaN,
                  hyperparams = hyperparams,
                  privacy_params = privacy_params,
                  swaps = 1, jumps = 1, fullUpdate = TRUE)
```

# Reduced run output

The output below reports the privatized sample size `N_DP`, a summary of the sample-size chain, and the first five saved draws of the regression-coefficient chain from this reduced run.

```{r}
chain$N_DP
summary(chain$n_chain)
round(chain$beta_chain[, 196:200], 3)
```

# Diagnostic figures

The figures below show simple trace plots for the regression coefficients and the sample-size chain in this reduced example.

```{r fig.width=5, fig.height=5}
matplot(t(chain$beta_chain), type = "l", ylim = c(-3, 3), xlab = "iteration", ylab = "beta chain")
abline(h = 0, col = "black")
abline(h = -1, col = "red")
abline(h = 1, col = "green")
```

```{r fig.width=5, fig.height=5}
plot(chain$n_chain, type = "l", xlab = "iteration", ylab = "N chain")
abline(h = 1000)
```

# Relation to the full repository

This notebook directly mirrors a reduced version of the Table 1 code path. Use the folders below for the full analyses reported in the paper:

- `Section6.1/Table1/` contains the full simulation and processing scripts for Table 1.
- `Section6.1/Table2/` and `Section6.1/Table3/` contain the other Section 6.1 simulation workflows.
- `Section6.2/` contains the ATUS application and plotting code.

Use those directories for the full analyses reported in the paper.
